Modern type-safe embedding of attribute grammars

### 2018-11-20 doc/Paper.tex
Update PDF to d22bf7b3c88c727408f851664c85bea5ffb5f91a

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\TODO{What's the title?}}

% TODO(twesterhout): This looks ugly... Someone, reformat it, please :)
\author{Jo{\~a}o Paulo Fernandes\inst{1}%
   \and Pedro Martins\inst{2}%
   \and Alberto Pardo\inst{3}%
   \and Jo{\~a}o Saraiva\inst{4}%
   \and Marcos Viera\inst{3}%
   \and Tom Westerhout\inst{5}%
}
\institute{%
  LISP/Release -- Universidade da Beira Interior, Portugal \email{jpf@di.ubi.pt} \and
  University of California, Irvine, USA \email{pribeiro@uci.edu} \and
  Universidad de la  Rep\'{u}blica, Uruguay \email{\{pardo,mviera\}@fing.edu.uy} \and
  Universidade do Minho, Portugal \email{saraiva@di.uminho.pt} \and
  Radboud University, The Netherlands \email{twesterhout@student.ru.nl}%
}

@@

\begin{abstract}
  \TODO{What's the abstract?}

@@

@@

  \paragraph{Lists} Suppose that we would like to update a list at a specific
  position:
\begin{tabbing}\texfamily
~~modify~::~(a~\WithMath{->}{\rightarrow}~[a])~\WithMath{->}{\rightarrow}~{\itshape Int}~\WithMath{->}{\rightarrow}~[a]~\WithMath{->}{\rightarrow}~[a]
\\
\texfamily ~~modify~f~i~xs~=~helper~[]~xs~0
\\
\texfamily ~~~{\bfseries where}~helper~before~(x~:~after)~!j
\\
\texfamily ~~~~~~~~~~~|~j~\WithMath{==}{\equiv}~i~~~~=~before~\WithMath{++}{+\!\!+}~f~x~\WithMath{++}{+\!\!+}~after
\\
\texfamily ~~~~~~~~~~~|~\textit{otherwise}~=~helper~(before~\WithMath{++}{+\!\!+}~[x])~after~(j~+~1)
\\
\texfamily ~~~~~~~~~helper~\char95 ~[]~\char95 ~=~\textit{error}~\char34 Index~out~of~bounds.\char34 
\end{tabbing}
  Here \text{\texfamily modify} takes an update action \text{\texfamily f}\footnote{\text{\texfamily f} returns a list rather
  than a single element to prevent curious readers from suggesting to use a
  boxed array instead of a list.}, an index \text{\texfamily i}, and a list \text{\texfamily xs} and returns a
  new list with the \text{\texfamily i}'th element replaced with the result of \text{\texfamily f}.%
  First, we ``unpack'' the list into $\text{\texfamily before\Sp =\Sp [}\ttsub{\text{\texfamily xs}}{\text{\texfamily 0}},\dots
  \ttsub{\text{\texfamily xs}}{\text{\texfamily i-1}}\text{\texfamily ]}$, $\text{\texfamily x\Sp =\Sp }\ttsub{\text{\texfamily xs}}{\text{\texfamily i}}$, and
  $\text{\texfamily after\Sp =\Sp [}\ttsub{\text{\texfamily xs}}{\text{\texfamily i+1}},\dots\text{\texfamily ]}$. Then replace \text{\texfamily x} by \text{\texfamily f(x)} and
  finally ``pack'' the result back into a single list. If we do a lot of
  updates, we end up unpacking and packing the list over and over again -- very
  time-consuming for long lists. We would thus like to benefit from fusion
  without explicitly working with the unpacked representation as it is
  bug-prone. A list zipper provides a way to achieve this.

  A zipper consists of a focus (alternatively called a hole) and surrounding
  context\footnote{`\text{\texfamily !}' in front of the type of \text{\texfamily \char95 cxt} is a
  strictness annotation enabled by the \text{\ttfamily BangPatterns} extension. Context can be
  seen as a path from the to the current focus and is always finite, i.e. there
  is no reason for it to be lazy.}:
\begin{tabbing}\texfamily
~~{\bfseries data}~{\itshape Zipper}~a~~=~{\itshape Zipper}~\char123 ~\char95 hole~::~a,~\char95 cxt~::~!({\itshape Context}~a)~\char125 
\\
\texfamily ~~{\bfseries data}~{\itshape Context}~a~=~{\itshape Context}~[a]~[a]
\end{tabbing}
  where the \text{\texfamily {\itshape Context}} keeps track of elements to the left (\text{\texfamily before} in
  \text{\texfamily modify}) and to the right (\text{\texfamily after} in \text{\texfamily modify}) of the focus. We can now
  define movements:
\begin{tabbing}\texfamily
~~left~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)
\\
\texfamily ~~left~({\itshape Zipper}~\char95 ~~~~({\itshape Context}~[]~\char95 ))~~~~~~~~=~{\itshape Nothing}
\\
\texfamily ~~left~({\itshape Zipper}~hole~({\itshape Context}~(l~:~ls)~rs))~=~{\itshape Just}~\$
\\
\texfamily ~~~~{\itshape Zipper}~l~\$~{\itshape Context}~ls~(hole~:~rs)
\\
\texfamily 
\\
\texfamily ~~right~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)
\\
\texfamily ~~right~({\itshape Zipper}~\char95 ~~~~({\itshape Context}~\char95 ~[]))~~~~~~~~=~{\itshape Nothing}
\\
\texfamily ~~right~({\itshape Zipper}~hole~({\itshape Context}~ls~(r~:~rs)))~=~{\itshape Just}~\$
\\
\texfamily ~~~~{\itshape Zipper}~r~\$~{\itshape Context}~(hole~:~ls)~rs
\end{tabbing}
  and functions for entering and leaving the zipper:
\begin{tabbing}\texfamily
~~enter~::~[a]~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)
\\
\texfamily ~~enter~[]~~~~~~~=~{\itshape Nothing}
\\
\texfamily ~~enter~(x~:~xs)~=~{\itshape Just}~\$~{\itshape Zipper}~x~({\itshape Context}~[]~xs)
\\
\texfamily 
\\
\texfamily ~~leave~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~[a]
\\
\texfamily ~~leave~({\itshape Zipper}~hole~({\itshape Context}~ls~rs))~=~reverse~ls~\WithMath{++}{+\!\!+}~hole~:~rs
\end{tabbing}

  Finally, we define a local version of our \text{\texfamily modify} function (\TODO{Boy, is
  this function ugly...})
\begin{tabbing}\texfamily
~~modify~::~(a~\WithMath{->}{\rightarrow}~[a])~\WithMath{->}{\rightarrow}~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)
\\
\texfamily ~~modify~f~({\itshape Zipper}~hole~({\itshape Context}~ls~rs))~=~{\bfseries case}~f~hole~{\bfseries of}
\\
\texfamily ~~~~(x~:~xs)~\WithMath{->}{\rightarrow}~{\itshape Just}~\$~{\itshape Zipper}~x~({\itshape Context}~ls~(xs~\WithMath{++}{+\!\!+}~rs))
\\
\texfamily ~~~~[]~~~~~~~\WithMath{->}{\rightarrow}~{\bfseries case}~rs~{\bfseries of}
\\
\texfamily ~~~~~~(r~:~rs')~\WithMath{->}{\rightarrow}~{\itshape Just}~\$~{\itshape Zipper}~r~({\itshape Context}~ls~rs')
\\
\texfamily ~~~~~~[]~~~~~~~~\WithMath{->}{\rightarrow}~{\bfseries case}~ls~{\bfseries of}
\\
\texfamily ~~~~~~~~(l~:~ls')~\WithMath{->}{\rightarrow}~{\itshape Just}~\$~{\itshape Zipper}~l~({\itshape Context}~ls'~rs)
\\
\texfamily ~~~~~~~~[]~~~~~~~~\WithMath{->}{\rightarrow}~{\itshape Nothing}
\end{tabbing}
  using which we can perform multiple updates efficiently and with minimal code
  bloat\footnote{%
  Operator \text{\ttfamily \char62{}\char61{}\char62{}} comes from \text{\texfamily {\itshape {\itshape Control}.Monad}} module in \text{\ttfamily base}
  and has the following signature:%
\begin{tabbing}\texfamily
~~(>=>)~::~{\itshape Monad}~m~=>~(a~\WithMath{->}{\rightarrow}~m~b)~\WithMath{->}{\rightarrow}~(b~\WithMath{->}{\rightarrow}~m~c)~\WithMath{->}{\rightarrow}~a~\WithMath{->}{\rightarrow}~m~c
\end{tabbing}
%
  }:
\begin{tabbing}\texfamily
~~modifyExample~::~{\itshape IO}~()
\\
\texfamily ~~modifyExample~=~print~\$
\\
\texfamily ~~~~enter~@{\itshape Int}~>=>~right
\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~right
\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~modify~(const~[])
\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~modify~(return~\WithMath{.}{\circ}~(+1))
\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~left
\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~modify~(return~\WithMath{.}{\circ}~negate)
\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~return~\WithMath{.}{\circ}~leave~\$~[1,~2,~3,~4,~5]
\end{tabbing}

  \paragraph{Trees} Consider now a binary tree data structure:
\begin{tabbing}\texfamily
~~{\bfseries data}~{\itshape Tree}~a~=~{\itshape Leaf}~!a~|~{\itshape Fork}~({\itshape Tree}~a)~({\itshape Tree}~a)
\end{tabbing}
  A binary tree zipper is slightly more insteresting than the list zipper,
  because we can move up and down the tree as well as left and right. The zipper
  again consists of a hole (a subtree we are focused on) and its surrounding
  context (a path from the hole to the root of the tree):
\begin{tabbing}\texfamily
~~{\bfseries data}~{\itshape Zipper}~a~~=~{\itshape Zipper}~\char123 ~\char95 hole~::~{\itshape Tree}~a,~\char95 cxt~::~!({\itshape Context}~a)~\char125 
\\
\texfamily ~~{\bfseries data}~{\itshape Context}~a~=~{\itshape Top}
\\
\texfamily ~~~~~~~~~~~~~~~~~|~{\itshape Left}~!({\itshape Context}~a)~({\itshape Tree}~a)
\\
\texfamily ~~~~~~~~~~~~~~~~~|~{\itshape Right}~({\itshape Tree}~a)~!({\itshape Context}~a)
\end{tabbing}
  To move the zipper down, we ``unpack'' the current hole:
\begin{tabbing}\texfamily
~~down~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)
\\
\texfamily ~~down~({\itshape Zipper}~({\itshape Leaf}~\char95 )~~~\char95 )~~~=~{\itshape Nothing}
\\
\texfamily ~~down~({\itshape Zipper}~({\itshape Fork}~l~r)~cxt)~=~{\itshape Just}~\$~{\itshape Zipper}~r~({\itshape Right}~l~cxt)
\end{tabbing}
  \text{\texfamily {\itshape Context}} stores everything we need to reconstruct the hole, and \text{\texfamily up} does
  exactly that:
\begin{tabbing}\texfamily
~~up~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)
\\
\texfamily ~~up~({\itshape Zipper}~\char95 ~{\itshape Top})~=~{\itshape Nothing}
\\
\texfamily ~~up~({\itshape Zipper}~l~({\itshape Left}~cxt~r))~=~{\itshape Just}~\$~{\itshape Zipper}~({\itshape Fork}~l~r)~cxt
\\
\texfamily ~~up~({\itshape Zipper}~r~({\itshape Right}~l~cxt))~=~{\itshape Just}~\$~{\itshape Zipper}~({\itshape Fork}~l~r)~cxt
\end{tabbing}
  Implementations of \text{\texfamily left}, \text{\texfamily right}, and \text{\texfamily enter} are very similar to the
  list zipper case and are left as an exercise for the reader. \text{\texfamily leave} differs
  slightly in that we now move all the way up rather than left:
\begin{tabbing}\texfamily
~~leave~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Tree}~a
\\
\texfamily ~~leave~z~=~{\bfseries case}~up~z~{\bfseries of}
\\
\texfamily ~~~~{\itshape Just}~z'~\WithMath{->}{\rightarrow}~leave~z'
\\
\texfamily ~~~~{\itshape Nothing}~\WithMath{->}{\rightarrow}~\char95 hole~z
\end{tabbing}

@@

  The generic zipper we will use is very similar to the one presented
  in\cite{adams2010syz}. The most common technique in Haskell for supporting
  heterogeneous types is Existential Quantification. However, not every type can
  act as a hole. To support moving down the tree, we need the hole to be
  \textit{dissectible}, i.e. we would like to be able to dissect the value into
  the constructor and its arguments. Even though \text{\texfamily {\itshape Data}.{\itshape Data}.gfoldl} allows us to
  achieve this, we define out own typeclass which additionally allows us to
  propagate down arbitrary constraints:
\begin{tabbing}\texfamily
~~{\bfseries class}~{\itshape Dissectible}~(c~::~{\itshape Type}~\WithMath{->}{\rightarrow}~{\itshape Constraint})~(a~::~{\itshape Type})~{\bfseries where}
\\
\texfamily ~~~~dissect~::~a~\WithMath{->}{\rightarrow}~{\itshape Left}~c~a
\\
\texfamily 
\\
\texfamily ~~{\bfseries data}~{\itshape Left}~c~expects~{\bfseries where}
\\
\texfamily ~~~~{\itshape LOne}~~::~b~\WithMath{->}{\rightarrow}~{\itshape Left}~c~b
\\
\texfamily ~~~~{\itshape LCons}~::~(c~b,~{\itshape Dissectible}~c~b)
\\
\texfamily ~~~~~~~~~~=>~{\itshape Left}~c~(b~\WithMath{->}{\rightarrow}~expects)~\WithMath{->}{\rightarrow}~b~\WithMath{->}{\rightarrow}~{\itshape Left}~c~expects
\end{tabbing}
  For example, here is how we can make \text{\texfamily {\itshape Tree}} an instance of \text{\texfamily {\itshape Disssectible}}
\begin{tabbing}\texfamily
~~{\bfseries instance}~c~({\itshape Tree}~a)~=>~{\itshape Dissectible}~c~({\itshape Tree}~a)~{\bfseries where}
\\
\texfamily ~~~~dissect~({\itshape Fork}~l~r)~=~{\itshape LOne}~{\itshape Fork}~`{\itshape LCons}`~l~`{\itshape LCons}`~r
\\
\texfamily ~~~~dissect~x~=~{\itshape LOne}~x
\end{tabbing}
  We can unpack a \text{\texfamily {\itshape Fork}} and the zipper will thus be able to go down. \text{\texfamily {\itshape Leaf}}s,
  however, are left untouched and trying to go down from a \text{\texfamily {\itshape Leaf}} will return
  \text{\texfamily {\itshape Nothing}}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Modern type-safe embedding of attribute grammars}
\subtitle{An exercise with functional zippers in Haskell}

% TODO(twesterhout): This looks ugly... Someone, reformat it, please :)
\author{Jo{\~a}o Paulo Fernandes\inst{1}%
   \and Pedro Martins\inst{2}%
   \and Alberto Pardo\inst{3}%
   \and Jo{\~a}o Saraiva\inst{4}%
   \and Marcos Viera\inst{3}%
   \and Tom Westerhout\inst{5}%
}
\institute{%
  CISUC -- Universidade de Coimbra, Portugal \email{jpf@dei.uc.pt} \and
  University of California, Irvine, USA \email{pribeiro@uci.edu} \and
  Universidad de la  Rep\'{u}blica, Uruguay, \email{\{pardo,mviera\}@fing.edu.uy} \and
  Universidade do Minho, Portugal, \email{saraiva@di.uminho.pt} \and
  Radboud University, The Netherlands, \email{twesterhout@student.ru.nl}%
}

@@

\begin{abstract}

  % Introduction. In one sentence, what’s the topic?

  Attribute grammars are a powerful, declarative formalism to implement and
  reason about programs which, by design, are conveniently modular.

  % State the problem you tackle

  Although a full attribute grammar compiler can be tailored to specific needs,
  its implementation is highly non trivial, and its long term maintenance is a
  major endeavor.

  % Summarize (in one sentence) why nobody else has adequately answered the
  % research question yet.

  In fact, maintaining a traditional attribute grammar system is such a hard
  effort that most such system that were proposed in the past are no longer
  active.

  % Explain, in one sentence, how you tackled the research question.

  Our approach to implement attribute grammars is to write them as first class
  citizens of a modern functional programming language.

  % In one sentence, how did you go about doing the research that follows from
  % your big idea.

  We improve a previous zipped-based attribute grammar embedding making it
  non-intrusive (i.e. no changes need to be made to the user-defined data types)
  and type-safe. On top of that, we achieve clearer syntax by using modern
  Haskell extensions.

  % As a single sentence, what’s the key impact of your research?

  We believe our embedding can be employed in practice to implement elegant,
  efficient and modular solutions to real life programming challenges.

@@

  Attribute Grammars (\ags) are a declarative formalism that was proposed by
  Knuth~\cite{Knuth68} in the late 60s and allows the implementation and
  reasoning about programs in a modular and convenient way. A concrete \ag
  relies on a context-free grammar to define the syntax of a language, and on
  attributes associated to the productions of the grammar to define the
  semantics of that language.
  %while adding \emph{attributes} to it so that it is also possible to define
  %its semantics.
  \ags\ have been used in practice to specify real programming languages, like
  for example \HASKELL~\cite{DijkstraFS09}, as well as powerful pretty printing
  algorithms \cite{SPS99}, deforestation techniques \cite{joao07pepm} and
  powerful type systems \cite{MiddelkoopDS10}.

  When programming with \ags, modularity is achieved due to the possibility of
  defining and using different aspects of computations as separate attributes.
  Attributes are distinct computation units, typically quite simple and modular,
  that can be combined into elaborated solutions to complex programming
  problems. They can also be analyzed, debugged and maintained independently
  which eases program development and evolution.

  \ags\ have proven to be particularly useful to specify computations over
  trees: given one tree, several \ag\ systems such
  as~\cite{syngen,uuag,lrc,silver} take specifications of which values, or
  attributes, need to be computed on the tree and perform these computations.
  The design and coding efforts put into the creation, improvement and
  maintenance of these \ag\ systems, however, is tremendous, which often is an
  obstacle to achieving the success they deserve.

  An increasingly popular alternative approach to the use of \ags\ relies on
  embedding them as first class citizens of general purpose programming
  languages~\cite{Oege00,DBLP:conf/sblp/MartinsFS13,erlangAGs,kiama,doaitse09icfp,balestrieri}.
  This avoids the burden of implementing a totally new language and associated
  system by hosting it in state-of-the-art programming languages. Following this
  approach one then exploits the modern constructions and infrastructure that
  are already provided by those languages and focus on the particularities of
  the domain specific language being developed.

@@

  \paragraph{Lists} Suppose that we would like to update a list at a specific
  position:
\begin{tabbing}\texfamily
~~modify~::~(a~\WithMath{->}{\rightarrow}~[a])~\WithMath{->}{\rightarrow}~{\itshape Int}~\WithMath{->}{\rightarrow}~[a]~\WithMath{->}{\rightarrow}~[a]\\
\texfamily ~~modify~f~i~xs~=~helper~[]~xs~0\\
\texfamily ~~~{\bfseries where}~helper~before~(x~:~after)~!j\\
\texfamily ~~~~~~~~~~~|~j~\WithMath{==}{\equiv}~i~~~~=~before~\WithMath{++}{+\!\!+}~f~x~\WithMath{++}{+\!\!+}~after\\
\texfamily ~~~~~~~~~~~|~\textit{otherwise}~=~helper~(before~\WithMath{++}{+\!\!+}~[x])~after~(j~+~1)\\
\texfamily ~~~~~~~~~helper~\char95 ~[]~\char95 ~=~\textit{error}~\char34 Index~out~of~bounds.\char34 
\end{tabbing}
  Here \text{\texfamily modify} takes an update action \text{\texfamily f}\footnote{\text{\texfamily f} returns a list rather
  than a single element to prevent curious readers from suggesting to use a
  boxed array instead of a list.}, an index \text{\texfamily i}, and a list \text{\texfamily xs} and returns a
  new list with the \text{\texfamily i}'th element replaced with the result of \text{\texfamily f}.%
  First, we ``unpack'' the list into $\text{\texfamily before\Sp =\Sp [}\ttsub{\text{\texfamily xs}}{\text{\texfamily 0}},\dots
  \ttsub{\text{\texfamily xs}}{\text{\texfamily i-1}}\text{\texfamily ]}$, $\text{\texfamily x\Sp =\Sp }\ttsub{\text{\texfamily xs}}{\text{\texfamily i}}$, and
  $\text{\texfamily after\Sp =\Sp [}\ttsub{\text{\texfamily xs}}{\text{\texfamily i+1}},\dots\text{\texfamily ]}$. Then replace \text{\texfamily x} by \text{\texfamily f(x)} and
  finally ``pack'' the result back into a single list. If we do a lot of
  updates, we end up unpacking and packing the list over and over again -- very
  time-consuming for long lists. We would thus like to benefit from fusion
  without explicitly working with the unpacked representation as it is
  bug-prone. A list zipper provides a way to achieve this.

  A zipper consists of a focus (alternatively called a hole) and surrounding
  context\footnote{`\text{\texfamily !}' in front of the type of \text{\texfamily \char95 cxt} is a
  strictness annotation enabled by the \text{\ttfamily BangPatterns} extension. Context can be
  seen as a path from the to the current focus and is always finite, i.e. there
  is no reason for it to be lazy.}:
\begin{tabbing}\texfamily
~~{\bfseries data}~{\itshape Zipper}~a~~=~{\itshape Zipper}~\char123 ~\char95 hole~::~a,~\char95 cxt~::~!({\itshape Context}~a)~\char125 \\
\texfamily ~~{\bfseries data}~{\itshape Context}~a~=~{\itshape Context}~[a]~[a]
\end{tabbing}
  where the \text{\texfamily {\itshape Context}} keeps track of elements to the left (\text{\texfamily before} in
  \text{\texfamily modify}) and to the right (\text{\texfamily after} in \text{\texfamily modify}) of the focus. We can now
  define movements:
\begin{tabbing}\texfamily
~~left~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)\\
\texfamily ~~left~({\itshape Zipper}~\char95 ~~~~({\itshape Context}~[]~\char95 ))~~~~~~~~=~{\itshape Nothing}\\
\texfamily ~~left~({\itshape Zipper}~hole~({\itshape Context}~(l~:~ls)~rs))~=~{\itshape Just}~\$\\
\texfamily ~~~~{\itshape Zipper}~l~\$~{\itshape Context}~ls~(hole~:~rs)\\
\texfamily \\
\texfamily ~~right~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)\\
\texfamily ~~right~({\itshape Zipper}~\char95 ~~~~({\itshape Context}~\char95 ~[]))~~~~~~~~=~{\itshape Nothing}\\
\texfamily ~~right~({\itshape Zipper}~hole~({\itshape Context}~ls~(r~:~rs)))~=~{\itshape Just}~\$\\
\texfamily ~~~~{\itshape Zipper}~r~\$~{\itshape Context}~(hole~:~ls)~rs
\end{tabbing}
  and functions for entering and leaving the zipper:
\begin{tabbing}\texfamily
~~enter~::~[a]~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)\\
\texfamily ~~enter~[]~~~~~~~=~{\itshape Nothing}\\
\texfamily ~~enter~(x~:~xs)~=~{\itshape Just}~\$~{\itshape Zipper}~x~({\itshape Context}~[]~xs)\\
\texfamily \\
\texfamily ~~leave~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~[a]\\
\texfamily ~~leave~({\itshape Zipper}~hole~({\itshape Context}~ls~rs))~=~reverse~ls~\WithMath{++}{+\!\!+}~hole~:~rs
\end{tabbing}

  Finally, we define a local version of our \text{\texfamily modify} function (\TODO{Boy, is
  this function ugly...})
\begin{tabbing}\texfamily
~~modify~::~(a~\WithMath{->}{\rightarrow}~[a])~\WithMath{->}{\rightarrow}~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)\\
\texfamily ~~modify~f~({\itshape Zipper}~hole~({\itshape Context}~ls~rs))~=~{\bfseries case}~f~hole~{\bfseries of}\\
\texfamily ~~~~(x~:~xs)~\WithMath{->}{\rightarrow}~{\itshape Just}~\$~{\itshape Zipper}~x~({\itshape Context}~ls~(xs~\WithMath{++}{+\!\!+}~rs))\\
\texfamily ~~~~[]~~~~~~~\WithMath{->}{\rightarrow}~{\bfseries case}~rs~{\bfseries of}\\
\texfamily ~~~~~~(r~:~rs')~\WithMath{->}{\rightarrow}~{\itshape Just}~\$~{\itshape Zipper}~r~({\itshape Context}~ls~rs')\\
\texfamily ~~~~~~[]~~~~~~~~\WithMath{->}{\rightarrow}~{\bfseries case}~ls~{\bfseries of}\\
\texfamily ~~~~~~~~(l~:~ls')~\WithMath{->}{\rightarrow}~{\itshape Just}~\$~{\itshape Zipper}~l~({\itshape Context}~ls'~rs)\\
\texfamily ~~~~~~~~[]~~~~~~~~\WithMath{->}{\rightarrow}~{\itshape Nothing}
\end{tabbing}
  using which we can perform multiple updates efficiently and with minimal code
  bloat\footnote{%
  Operator \text{\ttfamily \char62{}\char61{}\char62{}} comes from \text{\texfamily {\itshape {\itshape Control}.Monad}} module in \text{\ttfamily base}
  and has the following signature:%
\begin{tabbing}\texfamily
~~(>=>)~::~{\itshape Monad}~m~=>~(a~\WithMath{->}{\rightarrow}~m~b)~\WithMath{->}{\rightarrow}~(b~\WithMath{->}{\rightarrow}~m~c)~\WithMath{->}{\rightarrow}~a~\WithMath{->}{\rightarrow}~m~c
\end{tabbing}
%
  }:
\begin{tabbing}\texfamily
~~modifyExample~::~{\itshape IO}~()\\
\texfamily ~~modifyExample~=~print~\$\\
\texfamily ~~~~enter~@{\itshape Int}~>=>~right\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~right\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~modify~(const~[])\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~modify~(return~\WithMath{.}{\circ}~(+1))\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~left\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~modify~(return~\WithMath{.}{\circ}~negate)\\
\texfamily ~~~~~~~~~~~~~~~~~>=>~return~\WithMath{.}{\circ}~leave~\$~[1,~2,~3,~4,~5]
\end{tabbing}

  \paragraph{Trees} Consider now a binary tree data structure:
\begin{tabbing}\texfamily
~~{\bfseries data}~{\itshape Tree}~a~=~{\itshape Leaf}~!a~|~{\itshape Fork}~({\itshape Tree}~a)~({\itshape Tree}~a)
\end{tabbing}
  A binary tree zipper is slightly more insteresting than the list zipper,
  because we can move up and down the tree as well as left and right. The zipper
  again consists of a hole (a subtree we are focused on) and its surrounding
  context (a path from the hole to the root of the tree):
\begin{tabbing}\texfamily
~~{\bfseries data}~{\itshape Zipper}~a~~=~{\itshape Zipper}~\char123 ~\char95 hole~::~{\itshape Tree}~a,~\char95 cxt~::~!({\itshape Context}~a)~\char125 \\
\texfamily ~~{\bfseries data}~{\itshape Context}~a~=~{\itshape Top}\\
\texfamily ~~~~~~~~~~~~~~~~~|~{\itshape Left}~!({\itshape Context}~a)~({\itshape Tree}~a)\\
\texfamily ~~~~~~~~~~~~~~~~~|~{\itshape Right}~({\itshape Tree}~a)~!({\itshape Context}~a)
\end{tabbing}
  To move the zipper down, we ``unpack'' the current hole:
\begin{tabbing}\texfamily
~~down~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)\\
\texfamily ~~down~({\itshape Zipper}~({\itshape Leaf}~\char95 )~~~\char95 )~~~=~{\itshape Nothing}\\
\texfamily ~~down~({\itshape Zipper}~({\itshape Fork}~l~r)~cxt)~=~{\itshape Just}~\$~{\itshape Zipper}~r~({\itshape Right}~l~cxt)
\end{tabbing}
  \text{\texfamily {\itshape Context}} stores everything we need to reconstruct the hole, and \text{\texfamily up} does
  exactly that:
\begin{tabbing}\texfamily
~~up~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Maybe}~({\itshape Zipper}~a)\\
\texfamily ~~up~({\itshape Zipper}~\char95 ~{\itshape Top})~=~{\itshape Nothing}\\
\texfamily ~~up~({\itshape Zipper}~l~({\itshape Left}~cxt~r))~=~{\itshape Just}~\$~{\itshape Zipper}~({\itshape Fork}~l~r)~cxt\\
\texfamily ~~up~({\itshape Zipper}~r~({\itshape Right}~l~cxt))~=~{\itshape Just}~\$~{\itshape Zipper}~({\itshape Fork}~l~r)~cxt
\end{tabbing}
  Implementations of \text{\texfamily left}, \text{\texfamily right}, and \text{\texfamily enter} are very similar to the
  list zipper case and are left as an exercise for the reader. \text{\texfamily leave} differs
  slightly in that we now move all the way up rather than left:
\begin{tabbing}\texfamily
~~leave~::~{\itshape Zipper}~a~\WithMath{->}{\rightarrow}~{\itshape Tree}~a\\
\texfamily ~~leave~z~=~{\bfseries case}~up~z~{\bfseries of}\\
\texfamily ~~~~{\itshape Just}~z'~\WithMath{->}{\rightarrow}~leave~z'\\
\texfamily ~~~~{\itshape Nothing}~\WithMath{->}{\rightarrow}~\char95 hole~z
\end{tabbing}

@@

  The generic zipper we will use is very similar to the one presented
  in\cite{adams2010syz}. The most common technique in Haskell for supporting
  heterogeneous types is Existential Quantification. However, not every type can
  act as a hole. To support moving down the tree, we need the hole to be
  \textit{dissectible}, i.e. we would like to be able to dissect the value into
  the constructor and its arguments. Even though \text{\texfamily {\itshape Data}.{\itshape Data}.gfoldl} allows us to
  achieve this, we define out own typeclass which additionally allows us to
  propagate down arbitrary constraints:
\begin{tabbing}\texfamily
~~{\bfseries class}~{\itshape Dissectible}~(c~::~{\itshape Type}~\WithMath{->}{\rightarrow}~{\itshape Constraint})~(a~::~{\itshape Type})~{\bfseries where}\\
\texfamily ~~~~dissect~::~a~\WithMath{->}{\rightarrow}~{\itshape Left}~c~a\\
\texfamily \\
\texfamily ~~{\bfseries data}~{\itshape Left}~c~expects~{\bfseries where}\\
\texfamily ~~~~{\itshape LOne}~~::~b~\WithMath{->}{\rightarrow}~{\itshape Left}~c~b\\
\texfamily ~~~~{\itshape LCons}~::~(c~b,~{\itshape Dissectible}~c~b)\\
\texfamily ~~~~~~~~~~=>~{\itshape Left}~c~(b~\WithMath{->}{\rightarrow}~expects)~\WithMath{->}{\rightarrow}~b~\WithMath{->}{\rightarrow}~{\itshape Left}~c~expects
\end{tabbing}
  For example, here is how we can make \text{\texfamily {\itshape Tree}} an instance of \text{\texfamily {\itshape Disssectible}}
\begin{tabbing}\texfamily
~~{\bfseries instance}~c~({\itshape Tree}~a)~=>~{\itshape Dissectible}~c~({\itshape Tree}~a)~{\bfseries where}\\
\texfamily ~~~~dissect~({\itshape Fork}~l~r)~=~{\itshape LOne}~{\itshape Fork}~`{\itshape LCons}`~l~`{\itshape LCons}`~r\\
\texfamily ~~~~dissect~x~=~{\itshape LOne}~x
\end{tabbing}
  We can unpack a \text{\texfamily {\itshape Fork}} and the zipper will thus be able to go down. \text{\texfamily {\itshape Leaf}}s,
  however, are left untouched and trying to go down from a \text{\texfamily {\itshape Leaf}} will return
  \text{\texfamily {\itshape Nothing}}.
